A simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses

### 2016-09-18 Rmd/JTB_ms.Rmd
update notation

bibliography: ~/library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes

---

@@

The nutritional exchange between corals and *Symbiodinium* directly underlies the capacity of corals to build coral reef ecosystems, worth trillions of US Dollars annually [@Costanza:2014ex]. However, the complex symbiotic metabolism of corals is vulnerable to disruption by numerous anthropogenic environmental perturbations [@Baker:2015kp], jeopardizing their future persistence. In order to understand and predict coral responses to complex changes in the environment, a mechanistic understanding of how multiple interacting factors drive the individual and emergent physiology of both symbiotic partners is necessary. Such a task is well suited for theoretical modeling frameworks such as Dynamic Energy Budget (DEB) theory [@Kooijman:2010vd], although the complexity of such theory makes these efforts inaccessible to many biologists [@Jager:2013bj]. In order to bridge this gap, we present here a simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses that aims to mechanistically integrate the impacts of complex environmental change on the physiological ecology of reef corals.

@@

While the work of @Muller:2009io applying DEB formalism to this system represents the most significant theoretical contribution in coral symbiosis research to date, we aim to strengthen the role of theory and broaden its potential application in three primary ways:

@@

# Model description
In this model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Coral biomass, symbiont biomass, and prey biomass have fixed, but different, molar N:C ratios (Table 1). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the "parallel complementary" formulation of @Kooijman:2010vd to specify these fluxes. The two state variables in this dynamical system are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area (and surface area is assumed proportional to volume for corals (i.e., they are "V1-morphs" in DEB terminology [@Kooijman:2010vd])), biomass increases exponentially during growth. The rate of increase in coral biomass (i.e. growth) and the ratio of symbiont to host biomass are the (i.e. symbiosis dynamics) are the responses of interest of the system. Below we describe the formulation of each flux involved in producing these responses.

@@

\begin{equation} U_X = {{UX_m \cdot X} \over {X + X_{KX}}} \end{equation}

@@

\begin{equation} U_N = {{UN_m \cdot N} \over {N + X_{KN}}} \end{equation}

Coral biomass formation is specified by a parallel complementary SU that combines carbon and nitrogen to form biomass. In addition to the carbon and nitrogen acquired by direct uptake (Eqs. 1 and 2), the coral recycles a portion of the nitrogen liberated by biomass turnover ($r_{NR}$), and receives surplus fixed carbon shared by the symbiont ($\rho_C$), such that the total biomass formation is specified as:

\begin{equation} Q_R = \bigg({1 \over Q_{Rm}} + {1 \over {\rho_C + U_X}} + {1 \over {(U_N + n_{NX}U_X + r_{NR})n_{NR}^{-1}}} - {1 \over {{\rho_C + U_X} + (U_N + n_{NX}U_X + r_{NR})n_{NR}^{-1}}}\bigg)^{-1} \end{equation}

where $Q_{Rm}$ is the maximum specific growth rate. Finally, a fixed portion of coral biomass is turned over to pay maintenance costs:

\begin{equation} T_R = \gamma_R \end{equation}

\begin{equation} {dR \over dt} = Q_R - \gamma_R \end{equation}

where $\gamma_R$ is the parameter specifying the turnover rate.

@@

\begin{equation} \rho_N = U_N + n_{NX}U_X + r_{NR} - n_{NR}Q_R \end{equation}

Due to the inherent "leakiness" of the parallel complementary SU formulation, there will always be some nitrogen shared with the symbiont even when coral biomass formation is strongly nitrogen-limited. Likewise, there is always a non-zero rejection flux of carbon from the coral biomass SU, which is assumed to be lost to the environment.

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; Wangpraseurt et al., in prep.]. The ratio of internal scalar irradiance to external downwelling irradiance can take a theoretical maximum value of 3 for a "flat coral model" [@Marcelino:2013hz], and may be reduced below 1 at high symbiont densities due to self-shading. We used data from Wangpraseurt et al. (in prep) to empirically derive this ratio as a function of symbiont density (expressed as symbiont to host biomass ratio), and subsequently multiply this quantity by the external downwelling irradiance $L$ and the effective light-absorbing surface area of symbiont biomass $ds$ to specify the amount of light absorbed:

\begin{equation} U_L = 3 \cdot \exp(-4.905 \cdot S/R) \cdot L \cdot ds \end{equation}

@@

\begin{equation} UCP = UCP_p + UCP_a \end{equation}

In addition to the inputs CO~2~ specified in Eqs. 7, additional CO~2~ representing the metabolic production of CO~2~ in the host liberated from host biomass turnover (note: recylced C from symbiont was removed so that photodamage turnover does not make tons of C available), is made available to the photosynthesis SU. fixed carbon is produced by the photosynthesis SU according to:

\begin{equation} UCS = \bigg({1 \over UCS_m} + {1 \over {U_L \cdot n_{LC}}} + {1 \over {UCP + r_{CR}}} - {1 \over {U_L \cdot n_{LC} + UCP + r_{CR}}}\bigg)^{-1} \div ROS \end{equation}

where $UCS_m$ is the maximum specific rate of photosynthesis, and $ROS$ is the photooxidative stress multiplier (see below). Dividing by $ROS$ causes the rate of photosynthesis to decline in response to photo-oxidative stress, a phenomenon known as photoinhibition.

@@

\begin{equation} e_L = U_L - UCS \cdot n_{LC} \end{equation}

This excess light energy must be quenched by alternative pathways in order to prevent photo-oxidative damage. *Symbiodinium* may utilize a variety of pathways for non-photochemical quenching [NPQ; @Roth:2014wf], which we collect in a total capacity for NPQ as a parameter of the symbiont ($NPQ$). If light energy exceeds the total capacity of both carbon fixation and NPQ, then damaging reactive oxygen species (ROS) may be produced, which may have negative downstream consequences. We specify a generalized photo-oxidative stress flux $ROS$ which takes a value of 1 in the absence of stress, and increases as the amount of excess excitation energy increases.

\begin{equation} ROS = 1 + \bigg[\bigg({{e_L/S - NPQ} \over L50}\bigg)^k\bigg]_+ \end{equation}

where $NPQ$, $L50$, and $k$ are parameters of the symbiont that determine the overall susceptibility to photo-oxidative stress. Importantly, $ROS$ as specified here is not a function of absolute light absorption, but rather the amount of excess light energy $e_L$ after accounting for carbon fixation and NPQ. A direct consequence of this formulation is that carbon-limitation of photosynthesis can lead to photo-oxidative stress, a mechanism of biological importance [@Wooldridge:2009p7807] that was not captured by previous representations of photo-oxidative stress [@Eynaud:2011tv]. Moreover, this formulation allows functional diversity among symbiont types to be explored by changing the parameters $NPQ$, $L50$, and $k$.

Carbon fixed by photosynthesis ($UCS$) is then used in conjunction with nitrogen shared by the host ($\rho_N$) to build symbiont biomass, following the SU equation:

\begin{equation} Q_S = \bigg({1 \over Q_{Sm}} + {1 \over UCS} + {1 \over {(\rho_N + r_{NS})n_{NR}^{-1}}} - {1 \over {UCS + (\rho_N + r_{NS})n_{NR}^{-1}}}\bigg)^{-1} \end{equation}

@@

\begin{equation} rho_C = UCS - Q_S \end{equation}

@@

Finally, symbiont biomass is lost at a rate of

\begin{equation} T_S = \gamma_S \cdot 5 \cdot ROS \end{equation}

>>>>

bibliography: ~/library.bib
output: 
  rticles::elsevier_article:
    fig_caption: yes
header-includes:
  - \usepackage{tabularx}
---

@@

The nutritional exchange between corals and *Symbiodinium* directly underlies the capacity of corals to build coral reef ecosystems, worth trillions of US Dollars annually [@Costanza:2014ex]. However, the complex symbiotic metabolism of corals is vulnerable to disruption by numerous anthropogenic environmental perturbations, jeopardizing their future persistence. In order to understand and predict coral responses to complex changes in the environment, a mechanistic understanding of how multiple interacting factors drive the individual and emergent physiology of both symbiotic partners is necessary. Such a task is well suited for theoretical modeling frameworks such as Dynamic Energy Budget (DEB) theory [@Kooijman:2010vd], although the complexity of such theory makes these efforts inaccessible to many biologists [@Jager:2013bj]. In order to bridge this gap, we present here a simplified dynamic bioenergetic model for coral-*Symbiodinium* symbioses that aims to mechanistically integrate the impacts of complex environmental change on the physiological ecology of reef corals.

@@

While the work of @Muller:2009io applying DEB formalism to this system represents the most significant theoretical contribution in coral symbiosis research to date, we aim to strengthen the role of theory and broaden its potential application to corals in three primary ways:

@@

# Model description
In this model of coral-algal symbiosis, carbon and nitrogen are acquired by each partner and used to construct biomass. A graphical representation of the model is presented in Fig. 1. We use C-moles as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the amount of biomass containing 1 mole of Carbon atoms. Host biomass ($H$), symbiont biomass ($S$), and prey biomass ($X$) have fixed, but different, molar N:C ratios (Table 1). Carbon and nitrogen are combined to produce biomass by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the "parallel complementary" formulation of @Kooijman:2010vd to specify these fluxes. The two state variables in this dynamical system are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area (and surface area is assumed proportional to volume for corals (i.e., they are "V1-morphs" in DEB terminology [@Kooijman:2010vd])), biomass increases exponentially during growth. The rate of increase in coral biomass (i.e. growth) and the ratio of symbiont to host biomass are the (i.e. symbiosis dynamics) are the responses of interest of the system. Below we describe the formulation of each flux involved in producing these responses.

Table 1

| Symbol  | Description                            | Value  | Units              |
| ------- | -------------------------------------- | ------ | ------------------ |
| $n_{NH}$ | N:C molar ratio in host biomass | 0.19 | -- |
| $n_{NS}$ | N:C molar ratio in symbiont biomass | 0.2 | -- |
| $n_{NX}$ | N:C molar ratio in prey biomass | 0.13 | -- |
| $j_{HT}^0$ | Specific turnover rate of host biomass | 0.03 | $d^{-1}$ |
| $j_{ST}^0$ | Specific turnover rate of symbiont biomass | 0.03 | $d^{-1}$ |
| $\sigma_{NH}$ | Proportion host nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CH}$ | Proportion host carbon turnover recycled | 0.9 | -- |
| $\sigma_{NS}$ | Proportion symbiont nitrogen turnover recycled | 0.9 | -- |
| $\sigma_{CS}$ | Proportion symbiont carbon turnover recycled | 0.9 | -- |
| $j_{Xm}$ | Maximum specific feeding rate of host | 0.1292 | $molX \cdot CmolH^{-1} \cdot d^{-1}$ |
| $K_X$ | Half-saturation constant for prey uptake by host | 20e-6 | $molX \cdot L^{-1}$ | 
| $j_{Nm}$ | Maximum specific DIN uptake rate by host | 0.048 | $molN \cdot CmolH^{-1} \cdot d^{-1}$ |
| $K_N$ | Half-saturation constant for DIN uptake by host | 0.46e-6 | $molN \cdot L^{-1}$ | 
| $j_{CO_2}^p$ | Passive CO~2~ delivery to symbiont | 4.04e-3 | $molC \cdot CmolH^{-1} \cdot d^{-1}$ |
| $j_{CO_2}^a$ | Active CO~2~ delivery to symbiont | 0.32--18.0 | $molC \cdot CmolH^{-1} \cdot d^{-1}$ |
| $j_{HGm}$ | Maximum specific growth rate of host | 1 | $d^{-1}$ |
| $n_{LC}$ | Quantum yield of photosynthesis | 0.1 | $molC \cdot mol ph^{-1}$ |
| $\bar{a}^*$ | Effective light-absorbing cross-section of symbiont | 1.34 | $m^2 \cdot CmolS^{-1}$ |
| $j_{NPQ}$ | Non-photochemical quenching capacity of symbiont | 40 | $mol ph \cdot CmolS^{-1} \cdot d^{-1}$|
| $k_{ROS}$ | Excess photon energy that doubles ROS prod. | 40-80 | $mol ph \cdot CmolS^{-1} \cdot d^{-1}$|
| $k$ | Exponent on ROS production rate | 1 | -- |
| $j_{CPm}$ | Maximum specific photosynthesis rate of symbiont | 2.8 | $molC \cdot CmolS^{-1} \cdot d^{-1}$ |
| $j_{SGm}$ | Maximum specific growth rate of symbiont | 0.25 | $d^{-1}$ |

Also include tables of fluxes and environmental inputs?

@@

\begin{equation} j_X = {{j_{Xm} \cdot X} \over {X + K_X}} \end{equation}

@@

\begin{equation} j_N = {{j_{Nm} \cdot N} \over {N + K_N}} \end{equation}

Coral biomass formation is specified by a parallel complementary SU that combines carbon and nitrogen to form biomass. In addition to the carbon and nitrogen acquired by direct uptake (Eqs. 1 and 2), the coral recycles a portion of the nitrogen liberated by biomass turnover ($r_{NH}=\sigma_{NH}j_{HT}^0$), and receives surplus fixed carbon shared by the symbiont ($\rho_C$), such that the total biomass formation is specified as:

\begin{equation} j_{HG} = \bigg({1 \over j_{HGm}} + {1 \over {\rho_C{S \over H} + j_X}} + {1 \over {(j_N + n_{NX}j_X + r_{NH}) / n_{NH}}} - {1 \over {{\rho_C{S \over H} + j_X} + (j_N + n_{NX}j_X + r_{NH}) / n_{NH}}}\bigg)^{-1} \end{equation}

where $j_{HGm}$ is the maximum specific growth rate. Finally, the rate of change in coral biomass is calculated as the specific growth rate minum the specific biomass turnover rate (i.e., maintenance costs), according to:

\begin{equation} {dH \over dt} = j_{HG} - j_{HT}^0 \end{equation}

@@

\begin{equation} \rho_N = j_N + n_{NX}j_X + r_{NH} - n_{NH}j_{HG} \end{equation}

Due to the inherent inefficiency of the parallel complementary SU formulation, there will always be some nitrogen shared with the symbiont even when coral biomass formation is strongly nitrogen-limited. Likewise, there is always a non-zero rejection flux of carbon from the coral biomass SU, which is assumed to be lost to the environment.

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; @Marcelino:2013hz]. We used data from @Marcelino:2013hz to empirically derive the ratio of internal scalar irradiance to external downwelling irradiance as a function of symbiont density (expressed as symbiont to host biomass ratio), and subsequently multiply this quantity by the external downwelling irradiance $L$ and the effective light-absorbing surface area of symbiont biomass $ds$ to specify the amount of light absorbed:

\begin{equation} j_L = \big[1.26 + 1.39 \cdot \exp(-6.48 \cdot {S \over H})\big] \cdot L \cdot \bar{a}^* \end{equation}

@@

\begin{equation} j_{CO_2} = j_{CO_2}^p + j_{CO_2}^a \end{equation}

In addition to the inputs CO~2~ specified in Eqs. 7, additional CO~2~ representing the metabolic production of CO~2~ from host and symbiont biomass turnover ($r_{CH}=\sigma_{CH}j_{HT}^0; r_{CS}=\sigma_{CS}j_{ST}^0$) is made available to the photosynthesis SU. fixed carbon is produced by the photosynthesis SU according to:

\begin{equation} j_{CP} = \bigg({1 \over j_{CPm}} + {1 \over {n_{LC} j_L }} + {1 \over {(j_{CO_2} + r_{CH}){H \over S} + r_{CS}}} - {1 \over {n_{LC} j_L + (j_{CO_2} + r_{CH}){H \over S} + r_{CS}}}\bigg)^{-1} \div c_{ROS} \end{equation}

where $j_{CPm}$ is the maximum specific rate of photosynthesis, and $c_{ROS}$ is the photooxidative stress multiplier (see below). Dividing by $c_{ROS}$ causes the rate of photosynthesis to decline in response to photo-oxidative stress, a phenomenon known as photoinhibition.

@@

\begin{equation} j_{eL} = j_L - j_{CP} / n_{LC} \end{equation}

This excess light energy must be quenched by alternative pathways in order to prevent photo-oxidative damage. *Symbiodinium* may utilize a variety of pathways for non-photochemical quenching [NPQ; @Roth:2014wf], which we collect in a total capacity for NPQ as a parameter of the symbiont ($j_{NPQ}$). If light energy exceeds the total capacity of both carbon fixation and NPQ, then damaging reactive oxygen species (ROS) may be produced. We represent this as a scaled flux of ROS $c_{ROS}$, which takes a value of 1 when all light absorbed is quenched by photochemistry and NPQ, and increases as the amount of excess excitation energy increases.

\begin{equation} c_{ROS} = 1 + \bigg[\bigg({{j_{eL} - j_{NPQ}} \over k_{ROS}}\bigg)^k\bigg]_+ \end{equation}

where $j_{NPQ}$, $k_{ROS}$, and $k$ are parameters of the symbiont that determine the onset and rate of ROS production. Importantly, $c_{ROS}$ as specified here is not a function of absolute light absorption, but rather the amount of excess light energy $j_{eL}$ after accounting for carbon fixation and NPQ. A direct consequence of this formulation is that carbon-limitation of photosynthesis can lead to photo-oxidative stress, a mechanism of biological importance [@Wooldridge:2009p7807] that was not captured by previous representations of photo-oxidative stress [@Eynaud:2011tv]. Moreover, this formulation allows functional diversity among symbiont types to be explored by changing the parameters $j_{NPQ}$, $k_{ROS}$, and $k$.

Carbon fixed by photosynthesis ($j_{CP}$) is then used in conjunction with nitrogen shared by the host ($\rho_N$) and a proportion of nitrogen recycled from symbiont biomass turnover ($r_{NS}=\sigma_{NS}j_{ST}^0$) to build new symbiont biomass, following the SU equation:

\begin{equation} j_{SG} = \bigg({1 \over j_{SGm}} + {1 \over j_{CP}} + {1 \over {(\rho_N{H \over S} + r_{NS}) / n_{NH}}} - {1 \over {j_{CP} + (\rho_N{H \over S} + r_{NS}) / n_{NH}}}\bigg)^{-1} \end{equation}

@@

\begin{equation} \rho_C = j_{CP} - j_{SG} \end{equation}

@@

Symbiont biomass turnover includes a component of constant turnover specified by the parameter $j_{ST}^0$, representing fixed maintenance costs, plus a component that scales with the magnitude of ROS production.

\begin{equation} j_{ST} = j_{ST}^0(1 + 5 \cdot (c_{ROS}-1)) \end{equation}

This second component of symbiont biomass loss can represent both photodamage and/or symbiont expulsion (i.e., bleaching), both of which happen in response to higher levels or ROS production. The constant 5 is included to increase biomass loss in response to ROS. (Note that recycling of symbiont biomass turnover ($rNS$ and $rCS$) only occurs based on the basal maintenance related turnover (i.e., $j_{ST}^0$), and not the bleaching-related biomass loss, as this loss represents biomass being damaged or being expelled from the holobiont). Finally, the balance equation representing the change in symbiont biomass over time can be specified as the growth rate minus the turnover rate:

\begin{equation} {dS \over dt} = j_{SG} - j_{ST} \end{equation}

## Numerical analysis

The differential equations specified in 7 and 14 were solved using an implicit time-stepping method because of the implicit variables of surplus sharing. Euler method?
